---
title: "Group 2 - Checkpoint 1"
author: "Ashley Steinkrauss, Madeline Sullivan, & Roel Rodriguez"
date: "2024-04-10"
output: html_document
knitr::opts_chunk:
  echo: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

### Clear the workspace
```{r}
rm(list = ls()) # Clear environment
gc()            # Clear memory
cat("\f")       # Clear console

```

### Install and load in necessary packages
```{r}
library(tidyverse)
library(ggplot2)
library(devtools)
library(tidyr)
library(dplyr)
library(knitr)
library(reshape2)
library(plotly) # Interactive graphs

#install.packages("usmap")
library(usmap) # Plot geographic data
library(scales) # Aesthetics for geographic plot

```

### Load in .csv files
```{r}
# The following line of code starting with 'experiment_txt' will allow you to
# select the necessary .csv files from your computer to import (when you un-
# comment it)
# experiment_txt <- read.csv(file.choose(), header = T)

# Import .csv files
#Fruit2013 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Fruit Prices 2013.csv") # Edit to match your directory path
#Fruit2013 = Fruit2013[, 1:9] # Grabs first 9 columns (Some excels files had empty slots that were being read in weird)

#Fruit2016 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Fruit Prices 2016.csv") # Edit to match your directory path
#Fruit2016 = Fruit2016[, 1:9]

#Fruit2020 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Fruit Prices 2020.csv") # Edit to match your directory path
#Fruit2020 = Fruit2020[, 1:9]

#Veg2013 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Vegetable Prices 2013.csv") # Edit to match your directory path
#Veg2013 = Veg2013[, 1:9]

#Veg2016 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Vegetable Prices 2016.csv") # Edit to match your directory path
#Veg2016 = Veg2016[, 1:9]

#Veg2020 = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Vegetable Prices 2020.csv") # Edit to match your directory path
#Veg2020 = Veg2020[, 1:9]


#Income = read.csv("/Users/ashleysteinkrauss/Desktop/7900/GroupProj/Annual person income by county.csv")

# DELETE THE BELOW CODE FOR CHECK POINT 1 & UN-COMMENT THE PRECEDING LINES
Fruit2013 = read.csv("Fruit Prices 2013.csv") #edit these to match your directory path
Fruit2013 = Fruit2013[, 1:9] #grabbing first 9 columns, some excels had empty slots that were being read in weird
Fruit2016 = read.csv("Fruit Prices 2016.csv")
Fruit2016 = Fruit2016[, 1:9]
Fruit2020 = read.csv("Fruit Prices 2020.csv")
Fruit2020 = Fruit2020[, 1:9]

Veg2013 = read.csv("Vegetable Prices 2013.csv")
Veg2013 = Veg2013[, 1:9]
Veg2016 = read.csv("Vegetable Prices 2016.csv")
Veg2016 = Veg2016[, 1:9]
Veg2020 = read.csv("Vegetable Prices 2020.csv")
Veg2020 = Veg2020[, 1:9]


Income = read.csv("Annual person income by county.csv")

```

### Format data to have 'year' in long format
```{r}
Fruits = rbind(Fruit2013, Fruit2016, Fruit2020) # Make long format by combining all repeated columns
Veggies = rbind(Veg2013, Veg2016, Veg2020)

AnnualIncome <- pivot_longer(Income, cols = c("X2013", "X2016", "X2020"), names_to = "Year", values_to = "Value") # Combine the 3 year columns into 1 column

```

### Descriptive statistics
```{r}
# For fruits - Aggregating the cup equivalent price by type of fruit, year, and form of fruit
fruit_prices <- aggregate(CupEquivalentPrice ~ Fruit + Year + Form, data = Fruits, FUN = mean)

# For vegetables - Aggregating the cup equivalent price by type of fruit, year, and form of fruit
veggie_prices <- aggregate(CupEquivalentPrice ~ Vegetable + Year + Form, data = Veggies, FUN = mean)

# Converts 'Year' to a factor with specific levels
fruit_prices$Year <- factor(fruit_prices$Year, levels = c("2013", "2016", "2020"))
veggie_prices$Year <- factor(veggie_prices$Year, levels = c("2013", "2016", "2020"))

# Creates a summary statistic table for fruits
fruit_summary <- Fruits %>%
  group_by(Year) %>%
  summarise(mean_price = mean(CupEquivalentPrice),
            median_price = median(CupEquivalentPrice),
            min_price = min(CupEquivalentPrice),
            max_price = max(CupEquivalentPrice))

# Creates a summary statistic table for vegetables
veggie_summary <- Veggies %>%
  group_by(Year) %>%
  summarise(mean_price = mean(CupEquivalentPrice),
            median_price = median(CupEquivalentPrice),
            min_price = min(CupEquivalentPrice),
            max_price = max(CupEquivalentPrice))

# Views the fruit summary table
cat("Fruit Summary:\n")
kable(fruit_summary, align = "c")

# Views the vegetable summary table
cat("\nVeggie Summary:\n")
kable(veggie_summary, align = "c")

```

## Questions of interest

### 1. What are the most and least expensive fruits? How does this change across years?
```{r}
# Sort the Fruit2013 data frame by RetailPrice
sorted_Fruit2013 <- Fruit2013[order(Fruit2013$RetailPrice), ]
sorted_Fruit2016 <- Fruit2016[order(Fruit2016$RetailPrice), ]
sorted_Fruit2020 <- Fruit2020[order(Fruit2020$RetailPrice), ]

# Select the top 10 most expensive fruits
top_10_expensive_Fruit2013 <- tail(sorted_Fruit2013, 10)
top_10_expensive_Fruit2016 <- tail(sorted_Fruit2016, 10)
top_10_expensive_Fruit2020 <- tail(sorted_Fruit2020, 10)

# Select the bottom 10 least expensive fruits
bottom_10_expensive_Fruit2013 <- head(sorted_Fruit2013, 10)
bottom_10_expensive_Fruit2016 <- head(sorted_Fruit2016, 10)
bottom_10_expensive_Fruit2020 <- head(sorted_Fruit2020, 10)

# Create bar chart for most expensive fruits (sorted from least to most expensive)
most_expensive_Fruit2013 <- ggplot(top_10_expensive_Fruit2013, aes(x = RetailPrice, y = reorder(Fruit, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2013",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the first graph
print(most_expensive_Fruit2013)

# Create bar chart for least expensive fruits (sorted from most to least expensive)
least_expensive_Fruit2013 <- ggplot(bottom_10_expensive_Fruit2013, aes(x = RetailPrice, y = reorder(Fruit, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2013",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the second graph
print(least_expensive_Fruit2013)

# Create bar chart for most expensive fruits for 2016 (sorted from least to most expensive)
least_expensive_Fruit2016 <- ggplot(top_10_expensive_Fruit2016, aes(x = RetailPrice, y = reorder(Fruit, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2016",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the first graph for 2016
print(least_expensive_Fruit2016)

# Create bar chart for least expensive fruits for 2016 (sorted from most to least expensive)
least_expensive_Fruit2016 <- ggplot(bottom_10_expensive_Fruit2016, aes(x = RetailPrice, y = reorder(Fruit, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2016",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the second graph for 2016
print(least_expensive_Fruit2016)

# Create bar chart for most expensive fruits for 2020 (sorted from least to most expensive)
most_expensive_Fruit2020 <- ggplot(top_10_expensive_Fruit2020, aes(x = RetailPrice, y = reorder(Fruit, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2020",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the first graph for 2020
print(most_expensive_Fruit2020)

# Create bar chart for least expensive fruits for 2020 (sorted from most to least expensive)
least_expensive_Fruit2020 <- ggplot(bottom_10_expensive_Fruit2020, aes(x = RetailPrice, y = reorder(Fruit, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2020",
       x = "Retail Price",
       y = "Fruit") +
  theme_minimal()

# Print the second graph for 2020
print(least_expensive_Fruit2020)

```
```{r}
# Interactive version of the above graphs

# Create function to generate bar chart
create_bar_chart <- function(data, title, year, ascending = TRUE) {
  sorted_data <- data[order(data$RetailPrice, decreasing = !ascending), ]
  top_10_expensive <- tail(sorted_data, 10)
  bottom_10_expensive <- head(sorted_data, 10)
  
  most_expensive_plot <- ggplot(top_10_expensive, aes(x = RetailPrice, y = reorder(Fruit, RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
    labs(title = paste("Most Expensive", title, year),
         x = "Retail Price",
         y = "Fruit") +
    theme_minimal()
  
  least_expensive_plot <- ggplot(bottom_10_expensive, aes(x = RetailPrice, y = reorder(Fruit, -RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
    labs(title = paste("Least Expensive", title, year),
         x = "Retail Price",
         y = "Fruit") +
    theme_minimal()
  
  return(list(most_expensive_plot, least_expensive_plot))
}

# Create plots for each year
plots_2013 <- create_bar_chart(Fruit2013, "Fruits", "2013")
plots_2016 <- create_bar_chart(Fruit2016, "Fruits", "2016")
plots_2020 <- create_bar_chart(Fruit2020, "Fruits", "2020")

# Create plotly objects for most expensive fruits
plotly_most_expensive <- subplot(
  plotly::ggplotly(plots_2013[[1]]), 
  plotly::ggplotly(plots_2016[[1]]),
  plotly::ggplotly(plots_2020[[1]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Most Expensive Fruits")

# Create plotly objects for least expensive fruits
plotly_least_expensive <- subplot(
  plotly::ggplotly(plots_2013[[2]]), 
  plotly::ggplotly(plots_2016[[2]]),
  plotly::ggplotly(plots_2020[[2]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Least Expensive Fruits")

# Show the plotly objects
plotly_most_expensive
plotly_least_expensive

```

### 2. What are the most and least expensive vegetables? How does this change across years?
```{r}
# Sort the Veg2013 data frame by RetailPrice
sorted_Veg2013 <- Veg2013[order(Veg2013$RetailPrice), ]
sorted_Veg2016 <- Veg2016[order(Veg2016$RetailPrice), ]
sorted_Veg2020 <- Veg2020[order(Veg2020$RetailPrice), ]

# Select the top 10 most expensive Vegs
top_10_expensive_Veg2013 <- tail(sorted_Veg2013, 10)
top_10_expensive_Veg2016 <- tail(sorted_Veg2016, 10)
top_10_expensive_Veg2020 <- tail(sorted_Veg2020, 10)

# Select the bottom 10 least expensive Vegs
bottom_10_expensive_Veg2013 <- head(sorted_Veg2013, 10)
bottom_10_expensive_Veg2016 <- head(sorted_Veg2016, 10)
bottom_10_expensive_Veg2020 <- head(sorted_Veg2020, 10)

# Create bar chart for most expensive Vegs (sorted from least to most expensive)
most_expensive_Veg2013 <- ggplot(top_10_expensive_Veg2013, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegs 2013",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the first graph
print(most_expensive_Veg2013)

# Create bar chart for least expensive Vegs (sorted from most to least expensive)
least_expensive_Veg2013 <- ggplot(bottom_10_expensive_Veg2013, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegs 2013",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the second graph
print(least_expensive_Veg2013)

# Create bar chart for most expensive Vegs for 2016 (sorted from least to most expensive)
least_expensive_Veg2016 <- ggplot(top_10_expensive_Veg2016, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegs 2016",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the first graph for 2016
print(least_expensive_Veg2016)

# Create bar chart for least expensive Vegs for 2016 (sorted from most to least expensive)
least_expensive_Veg2016 <- ggplot(bottom_10_expensive_Veg2016, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegs 2016",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the second graph for 2016
print(least_expensive_Veg2016)

# Create bar chart for most expensive Vegs for 2020 (sorted from least to most expensive)
most_expensive_Veg2020 <- ggplot(top_10_expensive_Veg2020, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegs 2020",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the first graph for 2020
print(most_expensive_Veg2020)

# Create bar chart for least expensive Vegs for 2020 (sorted from most to least expensive)
least_expensive_Veg2020 <- ggplot(bottom_10_expensive_Veg2020, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegs 2020",
       x = "Retail Price",
       y = "Veg") +
  theme_minimal()

# Print the second graph for 2020
print(least_expensive_Veg2020)

```

```{r}
# Interactive version of the above graphs

# Create function to generate bar chart
create_bar_chart_veg <- function(data, title, year, ascending = TRUE) {
  sorted_data <- data[order(data$RetailPrice, decreasing = !ascending), ]
  top_10_expensive <- tail(sorted_data, 10)
  bottom_10_expensive <- head(sorted_data, 10)
  
  most_expensive_plot <- ggplot(top_10_expensive, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
    labs(title = paste("Most Expensive", title, year),
         x = "Retail Price",
         y = "Vegetable") +
    theme_minimal()
  
  least_expensive_plot <- ggplot(bottom_10_expensive, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
    labs(title = paste("Least Expensive", title, year),
         x = "Retail Price",
         y = "Vegetable") +
    theme_minimal()
  
  return(list(most_expensive_plot, least_expensive_plot))
}

# Create plots for each year
plots_2013_veg <- create_bar_chart_veg(Veg2013, "Vegetables", "2013")
plots_2016_veg <- create_bar_chart_veg(Veg2016, "Vegetables", "2016")
plots_2020_veg <- create_bar_chart_veg(Veg2020, "Vegetables", "2020")

# Create plotly objects for most expensive vegetables
plotly_most_expensive_veg <- subplot(
  plotly::ggplotly(plots_2013_veg[[1]]), 
  plotly::ggplotly(plots_2016_veg[[1]]),
  plotly::ggplotly(plots_2020_veg[[1]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Most Expensive Vegetables")

# Create plotly objects for least expensive vegetables
plotly_least_expensive_veg <- subplot(
  plotly::ggplotly(plots_2013_veg[[2]]), 
  plotly::ggplotly(plots_2016_veg[[2]]),
  plotly::ggplotly(plots_2020_veg[[2]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Least Expensive Vegetables")

# Show the plotly objects
plotly_most_expensive_veg
plotly_least_expensive_veg

```

### 3. How does income per capita differ by state across years?
```{r}
# Extracts row values associated with income per capita (The income dataset includes multiple
# measure of income, but we are only interested in working with income per capita)
income_by_capita <- subset(AnnualIncome, Description == "Per capita personal income (dollars) 2/")

# Renames 'X2013', 'X2016', and 'X2020' to '2013', '2016' and '2020' in the 'Year' column
income_by_capita <- income_by_capita %>% 
  mutate(Year = ifelse(Year == "X2013", "2013", 
                ifelse(Year == "X2016", "2016",
                ifelse(Year == "X2020", "2020", Year))))

# Converts 'Year' column to numeric
income_by_capita$Year <- as.numeric(income_by_capita$Year)

# Convert the 'Value' column to numeric (if needed)
income_by_capita$Value <- as.numeric(income_by_capita$Value)

# Aggregates the value by year and state
agg_income <- aggregate(Value ~ Year + GeoFIPS + GeoName, data = income_by_capita, FUN = mean)

# Format the Value column in personal_income -- IS THIS STILL NECESSARY?
#agg_income$Value <- comma(agg_income$Value)

# Filters data for each year
income_2013 <- agg_income[agg_income$Year == 2013, ]
income_Fruit2016 <- agg_income[agg_income$Year == 2016, ]
income_Fruit2020 <- agg_income[agg_income$Year == 2020, ]

# Generates the maps for each year & retrieves US state boundaries data
us_states <- us_map(regions = "states")

# YEAR = 2013, 2016, & 2020
# Merges data using state abbreviations in us_states and FIPS codes in agg_income
map_income <- merge(us_states, agg_income, by.x = "full", by.y = "GeoName")

# Creates a plot with ggplot
g <- ggplot(map_income) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income", palette = "Spectral") +
  ggtitle("Personal Income by State")

# Displays the ggplot graph
g

# Converts ggplot to plotly object
plotly_g <- ggplotly(g)

# Displays the interactive plot
plotly_g

# YEAR = 2013
# Merges data using state abbreviations in us_states and FIPS codes in agg_income
map_income_2013 <- merge(us_states, income_2013, by.x = "full", by.y = "GeoName")

# Creates a plot with ggplot
g_2013 <- ggplot(map_income_2013) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income", palette = "Spectral") +
  ggtitle("Personal Income by State in 2013")

# Displays the ggplot graph
g_2013

# Converts ggplot to plotly object
plotly_g_2013 <- ggplotly(g_2013)

# Displays the interactive plot
plotly_g_2013

# YEAR = 2016
# Merges data using state abbreviations in us_states and FIPS codes in agg_income
map_income_Fruit2016 <- merge(us_states, income_Fruit2016, by.x = "full", by.y = "GeoName")

# Creates a plot with ggplot
g_Fruit2016 <- ggplot(map_income_Fruit2016) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income", palette = "Spectral") +
  ggtitle("Personal Income by State in 2016")

# Displays the ggplot graph
g_Fruit2016

# Converts ggplot to plotly object
plotly_g_Fruit2016 <- ggplotly(g_Fruit2016)

# Displays the interactive plot
plotly_g_Fruit2016

# YEAR = 2020
# Merges using state abbreviations in us_states and FIPS codes in agg_income
map_income_Fruit2020 <- merge(us_states, income_Fruit2020, by.x = "full", by.y = "GeoName")
# Create a plot with ggplot
g_Fruit2020 <- ggplot(map_income_Fruit2020) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income", palette = "Spectral") +
  ggtitle("Personal Income by State in 2020")

# Displays the ggplot graph
g_Fruit2020

# Converts ggplot to plotly object
plotly_g_Fruit2020 <- ggplotly(g_Fruit2020)

# Displays the interactive plot
plotly_g_Fruit2020

```

### 4. How do the prices of different forms of fruits and vegetables change by year?
```{r}
# Hard codes years
Fruits$Year <- factor(Fruits$Year, levels = c("2013", "2016", "2020"))
Veggies$Year <- factor(Veggies$Year, levels = c("2013", "2016", "2020"))

# Aggregates the cup equivalent price by year, form of fruit, and type of fruit
avg_prices_fruit <- aggregate(CupEquivalentPrice ~ Year + Form + Fruit, data = Fruits, FUN = mean)
avg_prices_veg <- aggregate(CupEquivalentPrice ~ Year + Form + Vegetable, data = Veggies, FUN = mean)


# Creates a summary statistic table for fruits
fruit_summary_form <- Fruits %>%
  group_by(Year, Form) %>%
  summarise(mean_price = mean(CupEquivalentPrice),
            median_price = median(CupEquivalentPrice),
            min_price = min(CupEquivalentPrice),
            max_price = max(CupEquivalentPrice))

# Creates a summary statistic table for vegetables
veggie_summary_form <- Veggies %>%
  group_by(Year, Form) %>%
  summarise(mean_price = mean(CupEquivalentPrice),
            median_price = median(CupEquivalentPrice),
            min_price = min(CupEquivalentPrice),
            max_price = max(CupEquivalentPrice))

# Views the fruit summary
cat("Fruit Summary:\n")
kable(fruit_summary_form, align = "c")

# Views the vegetable summary
cat("\nVeggie Summary:\n")
kable(veggie_summary_form, align = "c")

# Converts ggplot graph to interactive plotly fruit graph
interactive_plot_fruit_form <- ggplotly(ggplot(avg_prices_fruit, aes(x = Year, y = CupEquivalentPrice, group = Fruit, color = Form)) +
                                geom_line() +
                                geom_point() +
                                labs(x = "Year", y = "Cup Equivalent Price",
                                     title = "Average Cup Equivalent Price of Fruits by Form Over Years", color = "Fruit Form") +
                                theme_minimal() +
                                theme(legend.position = "right"))

# Displays the interactive plot
interactive_plot_fruit_form

# Converts ggplot graph to interactive plotly vegetable graph
interactive_plot_veggie_form <- ggplotly(ggplot(avg_prices_veg, aes(x = Year, y = CupEquivalentPrice, group = Vegetable, color = Form)) +
                                geom_line() +
                                geom_point() +
                                labs(x = "Year", y = "Cup Equivalent Price",
                                     title = "Average Cup Equivalent Price of Fruits by Form Over Years", color = "Fruit Form") +
                                theme_minimal() +
                                theme(legend.position = "right"))

# Displays the interactive plot
interactive_plot_veggie_form

```

### 5. How many people can afford the recommended daily serving of fruits and vegetables across years?
```{r}
# Calculates 4% of per capita income
income_by_capita$eleven_percent <- income_by_capita$Value * 0.04

# Calculates 2% of per capita income
income_by_capita$four_percent =  income_by_capita$Value * 0.02

# Adjusts 'CupEquivalentPrice' to a yearly value by multiplying by 365
fruit_prices$CupEquivalentPrice_yearly <- fruit_prices$CupEquivalentPrice * 365
veggie_prices$CupEquivalentPrice_yearly <- veggie_prices$CupEquivalentPrice * 365

# Merges income_by_capita with fruit_prices by 'Year'
merged_fruits_2013 <- merge(subset(income_by_capita, Year == 2013), subset(fruit_prices, Year == 2013), by = "Year")
merged_fruits_Fruit2016 <- merge(subset(income_by_capita, Year == 2016), subset(fruit_prices, Year == 2016), by = "Year")
merged_fruits_Fruit2020 <- merge(subset(income_by_capita, Year == 2020), subset(fruit_prices, Year == 2020), by = "Year")

# Merges income_by_capita with veggie_prices by 'Year'
merged_veggies_2013 <- merge(subset(income_by_capita, Year == 2013), subset(veggie_prices, Year == 2013), by = "Year")
merged_veggies_Fruit2016 <- merge(subset(income_by_capita, Year == 2016), subset(veggie_prices, Year == 2016), by = "Year")
merged_veggies_Fruit2020 <- merge(subset(income_by_capita, Year == 2020), subset(veggie_prices, Year == 2020), by = "Year")

# Updates 'meetreq' column based on the condition for fruits
merged_fruits_2013$meetreq <- ifelse(merged_fruits_2013$four_percent > merged_fruits_2013$CupEquivalentPrice_yearly, "yes", "no")
merged_fruits_Fruit2016$meetreq <- ifelse(merged_fruits_Fruit2016$four_percent > merged_fruits_Fruit2016$CupEquivalentPrice_yearly, "yes", "no")
merged_fruits_Fruit2020$meetreq <- ifelse(merged_fruits_Fruit2020$four_percent > merged_fruits_Fruit2020$CupEquivalentPrice_yearly, "yes", "no")

# Updates 'meetreq' column based on the condition for vegetables
merged_veggies_2013$meetreq <- ifelse(merged_veggies_2013$four_percent > merged_veggies_2013$CupEquivalentPrice_yearly, "yes", "no")
merged_veggies_Fruit2016$meetreq <- ifelse(merged_veggies_Fruit2016$four_percent > merged_veggies_Fruit2016$CupEquivalentPrice_yearly, "yes", "no")
merged_veggies_Fruit2020$meetreq <- ifelse(merged_veggies_Fruit2020$four_percent > merged_veggies_Fruit2020$CupEquivalentPrice_yearly, "yes", "no")

# Creates comparison data frames for fruit in 2013
comparison_fruits_2013 <- data.frame(Year = merged_fruits_2013$Year,
                                     meetreq = merged_fruits_2013$meetreq,
                                     form = merged_fruits_2013$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_2013$CupEquivalentPrice_yearly)

# Creates comparison data frames for fruit in 2016
comparison_fruits_Fruit2016 <- data.frame(Year = merged_fruits_Fruit2016$Year,
                                     meetreq = merged_fruits_Fruit2016$meetreq,
                                     form = merged_fruits_Fruit2016$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_Fruit2016$CupEquivalentPrice_yearly)

# Creates comparison data frames for fruit in 2020
comparison_fruits_Fruit2020 <- data.frame(Year = merged_fruits_Fruit2020$Year,
                                     meetreq = merged_fruits_Fruit2020$meetreq,
                                     form = merged_fruits_Fruit2020$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_Fruit2020$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2013
comparison_veggies_2013 <- data.frame(Year = merged_veggies_2013$Year,
                                      meetreq = merged_veggies_2013$meetreq,
                                      form = merged_veggies_2013$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_2013$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2016
comparison_veggies_Fruit2016 <- data.frame(Year = merged_veggies_Fruit2016$Year,
                                      meetreq = merged_veggies_Fruit2016$meetreq,
                                      form = merged_veggies_Fruit2016$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_Fruit2016$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2020
comparison_veggies_Fruit2020 <- data.frame(Year = merged_veggies_Fruit2020$Year,
                                      meetreq = merged_veggies_Fruit2020$meetreq,
                                      form = merged_veggies_Fruit2020$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_Fruit2020$CupEquivalentPrice_yearly)


# Combines all comparison data frames for fruits and vegetables across different years
all_comparison_fruits <- rbind(comparison_fruits_2013, comparison_fruits_Fruit2016, comparison_fruits_Fruit2020)
all_comparison_veggies <- rbind(comparison_veggies_2013, comparison_veggies_Fruit2016, comparison_veggies_Fruit2020)

# Creates tables for fruits and vegetables
table_fruits <- table(all_comparison_fruits$meetreq, all_comparison_fruits$Year)
table_veggies <- table(all_comparison_veggies$meetreq, all_comparison_veggies$Year)

# Views fruit table
print("Fruits:")
print(table_fruits)

# Views vegetables table
print("Veggies:")
print(table_veggies)

```

### 6. How does the affordability of fruits and vegetables differ amongst their different forms?
```{r}
# Creates tables for fruits by form and year
table_fruits_form_2013 <- table(comparison_fruits_2013$meetreq, comparison_fruits_2013$form)
table_fruits_form_Fruit2016 <- table(comparison_fruits_Fruit2016$meetreq, comparison_fruits_Fruit2016$form)
table_fruits_form_Fruit2020 <- table(comparison_fruits_Fruit2020$meetreq, comparison_fruits_Fruit2020$form)

# Creatse tables for vegetables by form
table_veggies_form_2013 <- table(comparison_veggies_2013$meetreq, comparison_veggies_2013$form)
table_veggies_form_Fruit2016 <- table(comparison_veggies_Fruit2016$meetreq, comparison_veggies_Fruit2016$form)
table_veggies_form_Fruit2020 <- table(comparison_veggies_Fruit2020$meetreq, comparison_veggies_Fruit2020$form)

# Views 2013 fruit table
print("Fruits Summary by Form for 2013:")
print(table_fruits_form_2013)

# Views 2016 fruit table
print("Fruits Summary by Form for 2016:")
print(table_fruits_form_Fruit2016)
# Views 2020 fruit table
print("Fruits Summary by Form for 2020:")
print(table_fruits_form_Fruit2020)

# Views 2013 vegetable table
print("Veggies Summary by Form for 2013:")
print(table_veggies_form_2013)

# Views 2016 vegetable table
print("Veggies Summary by Form for 2016:")
print(table_veggies_form_Fruit2016)

# Views 2020 vegetable table
print("Veggies Summary by Form for 2020:")
print(table_veggies_form_Fruit2020)

```

### Further Analysis: Proportions of Affordabiltiy
```{r}
# Calculates proportions of yes/no affordability for fruits in 2013
table_fruits_form_2013_prop <- table_fruits_form_2013 / rowSums(table_fruits_form_2013)

# Calculates proportions of yes/no affordability for fruits in 2016
table_fruits_form_Fruit2016_prop <- table_fruits_form_Fruit2016 / rowSums(table_fruits_form_Fruit2016)

# Calculates proportions of yes/no affordability for fruits in 2020
table_fruits_form_Fruit2020_prop <- table_fruits_form_Fruit2020 / rowSums(table_fruits_form_Fruit2020)

# Calculates proportions of yes/no affordability for vegetables in 2013
table_veggies_form_2013_prop <- table_veggies_form_2013 / rowSums(table_veggies_form_2013)

# Calculates proportions of yes/no affordability for vegetables in 2016
table_veggies_form_Fruit2016_prop <- table_veggies_form_Fruit2016 / rowSums(table_veggies_form_Fruit2016)

# Calculates proportions of yes/no affordability for vegetables in 2020
table_veggies_form_Fruit2020_prop <- table_veggies_form_Fruit2020 / rowSums(table_veggies_form_Fruit2020)

```





