---
title: "Group 2"
author: "Ashley Steinkrauss, Madeline Sullivan, & Roel Rodriguez"
date: "2024-04-17"
output: html_document
always_allow_html: yes
knitr::opts_chunk:
  echo: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clear the workspace
```{r}
rm(list = ls()) # Clear environment
gc()            # Clear memory
cat("\f")       # Clear console
```

### Install and load in necessary packages
```{r}
#install.packages("usmap")
library(tidyverse)
library(ggplot2)
library(devtools)
library(tidyr)
library(dplyr)
library(stats)
library(knitr)
library(reshape2)
library(kableExtra)
library(plotly) # Interactive graphs
library(usmap) # Plot geographic data
library(scales) # Aesthetics for geographic plot
library(sf)
```

### Load in .csv files
```{r}
# The following line of code starting with 'experiment_txt' will allow you to
# select the necessary .csv files from your computer to import (when you un-
# comment it)
# experiment_txt <- read.csv(file.choose(), header = T)

Fruit2013 = read.csv("Data/Fruit Prices 2013.csv") #edit these to match your directory path
Fruit2013 = Fruit2013[, 1:9] #grabbing first 9 columns, some excels had empty slots that were being read in weird
Fruit2016 = read.csv("Data/Fruit Prices 2016.csv")
Fruit2016 = Fruit2016[, 1:9]
Fruit2020 = read.csv("Data/Fruit Prices 2020.csv")
Fruit2020 = Fruit2020[, 1:9]

Veg2013 = read.csv("Data/Vegetable Prices 2013.csv")
Veg2013 = Veg2013[, 1:9]
Veg2016 = read.csv("Data/Vegetable Prices 2016.csv")
Veg2016 = Veg2016[, 1:9]
Veg2020 = read.csv("Data/Vegetable Prices 2020.csv")
Veg2020 = Veg2020[, 1:9]

Income = read.csv("Data/Annual person income by county.csv")
```

### Format data to have 'year' in long format
```{r}
# Make long format by combining all repeated columns
Fruits = rbind(Fruit2013, Fruit2016, Fruit2020)
Veggies = rbind(Veg2013, Veg2016, Veg2020)

# Combine the 3 year columns into 1 column
AnnualIncome <- pivot_longer(Income, cols = c("X2013", "X2016", "X2020"), names_to = "Year", values_to = "Value")
```

### Descriptive statistics
```{r}
# For fruits - Aggregating the cup equivalent price by type of fruit, year, and form of fruit
fruit_prices <- aggregate(CupEquivalentPrice ~ Fruit + Year + Form, data = Fruits, FUN = mean)

# For vegetables - Aggregating the cup equivalent price by type of fruit, year, and form of fruit
veggie_prices <- aggregate(CupEquivalentPrice ~ Vegetable + Year + Form, data = Veggies, FUN = mean)

# Converts 'Year' to a factor with specific levels
fruit_prices$Year <- factor(fruit_prices$Year, levels = c("2013", "2016", "2020"))
veggie_prices$Year <- factor(veggie_prices$Year, levels = c("2013", "2016", "2020"))

# Creates a summary statistic table for fruits
fruit_summary <- Fruits %>%
  group_by(Year) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice))

# Creates a summary statistic table for vegetables
veggie_summary <- Veggies %>%
  group_by(Year) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice))
```

```{r}
# Aggregating the cup equivalent price by type of fruit and year for fruits
fruit_summary <- Fruits %>%
  group_by(Year) %>%
  summarise(Mean = mean(CupEquivalentPrice),
            Median = median(CupEquivalentPrice),
            Min = min(CupEquivalentPrice),
            Max = max(CupEquivalentPrice)) %>%
  mutate(Type = "Fruit")

# Aggregating the cup equivalent price by type of vegetable and year for vegetables
veggie_summary <- Veggies %>%
  group_by(Year) %>%
  summarise(Mean = mean(CupEquivalentPrice),
            Median = median(CupEquivalentPrice),
            Min = min(CupEquivalentPrice),
            Max = max(CupEquivalentPrice)) %>%
  mutate(Type = "Vegetable")

# Convert Year to factor with specific levels
fruit_summary$Year <- factor(fruit_summary$Year, levels = c("2013", "2016", "2020"))
veggie_summary$Year <- factor(veggie_summary$Year, levels = c("2013", "2016", "2020"))

# Creating pretty tables for descriptive statistics
fruit_table <- fruit_summary %>%
  select(Type, everything()) %>%
  kable(format = "markdown", caption = "Fruit Summary") %>%
  kable_styling()

veggie_table <- veggie_summary %>%
  select(Type, everything()) %>%
  kable(format = "markdown", caption = "Vegetable Summary") %>%
  kable_styling()

# View the tables
fruit_table
veggie_table
```

```{r}
fruit_data <- mutate(Fruits, Type = "Fruit")
veggie_data <- mutate(Veggies, Type = "Vegetable")

common_columns <- intersect(names(fruit_data), names(veggie_data))
fruit_data <- fruit_data[, common_columns]
veggie_data <- veggie_data[, common_columns]

# Combine fruit and vegetable data into one dataframe
data_combined <- rbind(fruit_data, veggie_data)

# Create a summary data frame with necessary statistics
summary_data <- data_combined %>%
  group_by(Year, Type) %>%
  summarise(
    mean_price = mean(CupEquivalentPrice, na.rm = TRUE)
  )

# Plotting
ggplot(summary_data, aes(x = factor(Year), y = mean_price, color = Type, group = Type)) +
  geom_line(size = 1) +
  geom_point(size = 3, shape = 21, fill = "white") +  # Adds points to the lines
  geom_text(aes(label = sprintf("$%.2f", mean_price)), vjust = -0.5, size = 3, color = "black") +
  labs(title = "Price Trends for Fruits and Vegetables",
       x = "Year",
       y = "Mean Cup Equivalent Price ($)") +
  scale_color_manual(values = c("skyblue", "lightgreen")) +
  theme_minimal()
```


```{r}
# Aggregating the cup equivalent price by type of fruit and year for fruits
fruit_summary <- Fruits %>%
  group_by(Year) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice)) %>%
  mutate(Type = "Fruit")

# Aggregating the cup equivalent price by type of vegetable and year for vegetables
veggie_summary <- Veggies %>%
  group_by(Year) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice)) %>%
  mutate(Type = "Vegetable")

# Convert Year to factor with specific levels
fruit_summary$Year <- factor(fruit_summary$Year, levels = c("2013", "2016", "2020"))
veggie_summary$Year <- factor(veggie_summary$Year, levels = c("2013", "2016", "2020"))

# Creating pretty tables for descriptive statistics
fruit_table <- fruit_summary %>%
  select(Type, everything()) %>%
  kable(format = "markdown", caption = "Fruit Summary") %>%
  kable_styling()

veggie_table <- veggie_summary %>%
  select(Type, everything()) %>%
  kable(format = "markdown", caption = "Vegetable Summary") %>%
  kable_styling()

# View the tables
fruit_table
veggie_table
```

## Questions of interest

### 1. What are the most and least expensive fruits? How does this change across years?
```{r}
# Sort the Fruit2013 data frame by RetailPrice
sorted_Fruit2013 <- Fruit2013[order(Fruit2013$RetailPrice), ]
sorted_Fruit2016 <- Fruit2016[order(Fruit2016$RetailPrice), ]
sorted_Fruit2020 <- Fruit2020[order(Fruit2020$RetailPrice), ]

# Select the top 10 most expensive fruits
top_10_expensive_Fruit2013 <- tail(sorted_Fruit2013, 10)
top_10_expensive_Fruit2016 <- tail(sorted_Fruit2016, 10)
top_10_expensive_Fruit2020 <- tail(sorted_Fruit2020, 10)

# Select the bottom 10 least expensive fruits
bottom_10_expensive_Fruit2013 <- head(sorted_Fruit2013, 10)
bottom_10_expensive_Fruit2016 <- head(sorted_Fruit2016, 10)
bottom_10_expensive_Fruit2020 <- head(sorted_Fruit2020, 10)

# Create bar chart for most expensive fruits (sorted from least to most expensive)
most_expensive_Fruit2013 <- ggplot(top_10_expensive_Fruit2013, aes(x = CupEquivalentPrice, y = reorder(Fruit, CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2013",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the first graph
print(most_expensive_Fruit2013)

# Create bar chart for least expensive fruits (sorted from most to least expensive)
least_expensive_Fruit2013 <- ggplot(bottom_10_expensive_Fruit2013, aes(x = CupEquivalentPrice, y = reorder(Fruit, -CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2013",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the second graph
print(least_expensive_Fruit2013)

# Create bar chart for most expensive fruits for 2016 (sorted from least to most expensive)
least_expensive_Fruit2016 <- ggplot(top_10_expensive_Fruit2016, aes(x = CupEquivalentPrice, y = reorder(Fruit, CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2016",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the first graph for 2016
print(least_expensive_Fruit2016)

# Create bar chart for least expensive fruits for 2016 (sorted from most to least expensive)
least_expensive_Fruit2016 <- ggplot(bottom_10_expensive_Fruit2016, aes(x = CupEquivalentPrice, y = reorder(Fruit, -CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2016",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the second graph for 2016
print(least_expensive_Fruit2016)

# Create bar chart for most expensive fruits for 2020 (sorted from least to most expensive)
most_expensive_Fruit2020 <- ggplot(top_10_expensive_Fruit2020, aes(x = CupEquivalentPrice, y = reorder(Fruit, CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Fruits 2020",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the first graph for 2020
print(most_expensive_Fruit2020)

# Create bar chart for least expensive fruits for 2020 (sorted from most to least expensive)
least_expensive_Fruit2020 <- ggplot(bottom_10_expensive_Fruit2020, aes(x = CupEquivalentPrice, y = reorder(Fruit, -CupEquivalentPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Fruits 2020",
       x = "Retail Price (per equivalent cup, $)",
       y = "Fruit") +
  theme_minimal()

# Print the second graph for 2020
print(least_expensive_Fruit2020)

```

```{r}
# Interactive version of the above graphs

# Create function to generate bar chart
create_bar_chart <- function(data, title, year, ascending = TRUE) {
  sorted_data <- data[order(data$RetailPrice, decreasing = !ascending), ]
  top_10_expensive <- tail(sorted_data, 10)
  bottom_10_expensive <- head(sorted_data, 10)
  
  most_expensive_plot <- ggplot(top_10_expensive, aes(x = RetailPrice, y = reorder(Fruit, RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
    labs(title = paste("Most Expensive", title, year),
         x = "Retail Price (per equivalent cup, $)",
         y = "Fruit") +
    theme_minimal()
  
  least_expensive_plot <- ggplot(bottom_10_expensive, aes(x = RetailPrice, y = reorder(Fruit, -RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
    labs(title = paste("Least Expensive", title, year),
         x = "Retail Price (per equivalent cup, $)",
         y = "Fruit") +
    theme_minimal()
  
  return(list(most_expensive_plot, least_expensive_plot))
}

# Create plots for each year
plots_2013 <- create_bar_chart(Fruit2013, "Fruits", "2013")
plots_2016 <- create_bar_chart(Fruit2016, "Fruits", "2016")
plots_2020 <- create_bar_chart(Fruit2020, "Fruits", "2020")

# Create plotly objects for most expensive fruits
plotly_most_expensive <- subplot(
  plotly::ggplotly(plots_2013[[1]]), 
  plotly::ggplotly(plots_2016[[1]]),
  plotly::ggplotly(plots_2020[[1]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Most Expensive Fruits")

# Create plotly objects for least expensive fruits
plotly_least_expensive <- subplot(
  plotly::ggplotly(plots_2013[[2]]), 
  plotly::ggplotly(plots_2016[[2]]),
  plotly::ggplotly(plots_2020[[2]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Least Expensive Fruits")

# Show the plotly objects
plotly_most_expensive
plotly_least_expensive

```

### 2. What are the most and least expensive vegetables? How does this change across years?
```{r}
# Sort the Veg2013 data frame by RetailPrice
sorted_Veg2013 <- Veg2013[order(Veg2013$RetailPrice), ]
sorted_Veg2016 <- Veg2016[order(Veg2016$RetailPrice), ]
sorted_Veg2020 <- Veg2020[order(Veg2020$RetailPrice), ]

# Select the top 10 most expensive Vegs
top_10_expensive_Veg2013 <- tail(sorted_Veg2013, 10)
top_10_expensive_Veg2016 <- tail(sorted_Veg2016, 10)
top_10_expensive_Veg2020 <- tail(sorted_Veg2020, 10)

# Select the bottom 10 least expensive Vegs
bottom_10_expensive_Veg2013 <- head(sorted_Veg2013, 10)
bottom_10_expensive_Veg2016 <- head(sorted_Veg2016, 10)
bottom_10_expensive_Veg2020 <- head(sorted_Veg2020, 10)

# Create bar chart for most expensive Vegs (sorted from least to most expensive)
most_expensive_Veg2013 <- ggplot(top_10_expensive_Veg2013, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegetables 2013",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the first graph
print(most_expensive_Veg2013)

# Create bar chart for least expensive Vegs (sorted from most to least expensive)
least_expensive_Veg2013 <- ggplot(bottom_10_expensive_Veg2013, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegetables 2013",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the second graph
print(least_expensive_Veg2013)

# Create bar chart for most expensive Vegs for 2016 (sorted from least to most expensive)
least_expensive_Veg2016 <- ggplot(top_10_expensive_Veg2016, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegetables 2016",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the first graph for 2016
print(least_expensive_Veg2016)

# Create bar chart for least expensive Vegs for 2016 (sorted from most to least expensive)
least_expensive_Veg2016 <- ggplot(bottom_10_expensive_Veg2016, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegetables 2016",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the second graph for 2016
print(least_expensive_Veg2016)

# Create bar chart for most expensive Vegs for 2020 (sorted from least to most expensive)
most_expensive_Veg2020 <- ggplot(top_10_expensive_Veg2020, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
  labs(title = "Most Expensive Vegetables 2020",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the first graph for 2020
print(most_expensive_Veg2020)

# Create bar chart for least expensive Vegs for 2020 (sorted from most to least expensive)
least_expensive_Veg2020 <- ggplot(bottom_10_expensive_Veg2020, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
  labs(title = "Least Expensive Vegetables 2020",
       x = "Retail Price (per equivalent cup, $)",
       y = "Vegetable") +
  theme_minimal()

# Print the second graph for 2020
print(least_expensive_Veg2020)

```

```{r}
# Interactive version of the above graphs

# Create function to generate bar chart
create_bar_chart_veg <- function(data, title, year, ascending = TRUE) {
  sorted_data <- data[order(data$RetailPrice, decreasing = !ascending), ]
  top_10_expensive <- tail(sorted_data, 10)
  bottom_10_expensive <- head(sorted_data, 10)
  
  most_expensive_plot <- ggplot(top_10_expensive, aes(x = RetailPrice, y = reorder(Vegetable, RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "skyblue") +
    labs(title = paste("Most Expensive", title, year),
         x = "Retail Price (per equivalent cup, $)",
         y = "Vegetable") +
    theme_minimal()
  
  least_expensive_plot <- ggplot(bottom_10_expensive, aes(x = RetailPrice, y = reorder(Vegetable, -RetailPrice))) +
    geom_bar(stat = "identity", position = "dodge", width = 0.7, fill = "lightgreen") +
    labs(title = paste("Least Expensive", title, year),
         x = "Retail Price (per equivalent cup, $)",
         y = "Vegetable") +
    theme_minimal()
  
  return(list(most_expensive_plot, least_expensive_plot))
}

# Create plots for each year
plots_2013_veg <- create_bar_chart_veg(Veg2013, "Vegetables", "2013")
plots_2016_veg <- create_bar_chart_veg(Veg2016, "Vegetables", "2016")
plots_2020_veg <- create_bar_chart_veg(Veg2020, "Vegetables", "2020")

# Create plotly objects for most expensive vegetables
plotly_most_expensive_veg <- subplot(
  plotly::ggplotly(plots_2013_veg[[1]]), 
  plotly::ggplotly(plots_2016_veg[[1]]),
  plotly::ggplotly(plots_2020_veg[[1]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Most Expensive Vegetables")

# Create plotly objects for least expensive vegetables
plotly_least_expensive_veg <- subplot(
  plotly::ggplotly(plots_2013_veg[[2]]), 
  plotly::ggplotly(plots_2016_veg[[2]]),
  plotly::ggplotly(plots_2020_veg[[2]]),
  nrows = 3, 
  margin = 0.05
) %>% layout(title = "Least Expensive Vegetables")

# Show the plotly objects
plotly_most_expensive_veg
plotly_least_expensive_veg

```

### 3. How does income per capita differ by state across years?
```{r}
# Extracts row values associated with income per capita (The income dataset includes multiple
# measure of income, but we are only interested in working with income per capita)
income_by_capita <- subset(AnnualIncome, Description == "Per capita personal income (dollars) 2/")

# Renames 'X2013', 'X2016', and 'X2020' to '2013', '2016' and '2020' in the 'Year' column
income_by_capita <- income_by_capita %>% 
  mutate(Year = ifelse(Year == "X2013", "2013", 
                ifelse(Year == "X2016", "2016",
                ifelse(Year == "X2020", "2020", Year))))

# Converts 'Year' column to numeric
income_by_capita$Year <- as.numeric(income_by_capita$Year)

# Convert the 'Value' column to numeric (if needed)
income_by_capita$Value <- as.numeric(income_by_capita$Value)

# Aggregates the value by year and state
agg_income <- aggregate(Value ~ Year + GeoFIPS + GeoName, data = income_by_capita, FUN = mean)

# Format the Value column in personal_income -- IS THIS STILL NECESSARY?
#agg_income$Value <- comma(agg_income$Value)

# Filters data for each year
income_2013 <- agg_income[agg_income$Year == 2013, ]
income_2016 <- agg_income[agg_income$Year == 2016, ]
income_2020 <- agg_income[agg_income$Year == 2020, ]
```
```{r}
# Generates the maps for each year & retrieves US state boundaries data
us_states <- us_map(regions = "states")

# YEAR = 2013
# Merges data using state abbreviations in us_states and FIPS codes in agg_income
map_income_2013 <- merge(us_states, income_2013, by.x = "full", by.y = "GeoName")

# Creates a plot with ggplot
g_2013 <- ggplot(map_income_2013) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income ($, by capita)", palette = "Spectral") +
  ggtitle("Personal Income by State (2013)")

# Displays the ggplot graph
g_2013

# Converts ggplot to plotly object
plotly_g_2013 <- ggplotly(g_2013)

# Displays the interactive plot
plotly_g_2013

# YEAR = 2016
# Merges data using state abbreviations in us_states and FIPS codes in agg_income
map_income_2016 <- merge(us_states, income_2016, by.x = "full", by.y = "GeoName")

# Creates a plot with ggplot
g_2016 <- ggplot(map_income_2016) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income ($, by capita)", palette = "Spectral") +
  ggtitle("Personal Income by State (2016)")

# Displays the ggplot graph
g_2016

# Converts ggplot to plotly object
plotly_g_2016 <- ggplotly(g_2016)

# Displays the interactive plot
plotly_g_2016

# YEAR = 2020
# Merges using state abbreviations in us_states and FIPS codes in agg_income
map_income_2020 <- merge(us_states, income_2020, by.x = "full", by.y = "GeoName")
# Create a plot with ggplot
g_2020 <- ggplot(map_income_2020) +
  geom_sf(aes(fill = Value)) +
  scale_fill_distiller(name = "Income ($, by capita)", palette = "Spectral") +
  ggtitle("Personal Income by State (2020)")

# Displays the ggplot graph
g_2020

# Converts ggplot to plotly object
plotly_g_2020 <- ggplotly(g_2020)

# Displays the interactive plot
plotly_g_2020
```

### 4. How do the prices of different forms of fruits and vegetables change by year?
```{r}
# Hard codes years
Fruits$Year <- factor(Fruits$Year, levels = c("2013", "2016", "2020"))
Veggies$Year <- factor(Veggies$Year, levels = c("2013", "2016", "2020"))

# Aggregates the cup equivalent price by year, form of fruit, and type of fruit
avg_prices_fruit <- aggregate(CupEquivalentPrice ~ Year + Form + Fruit, data = Fruits, FUN = mean)
avg_prices_veg <- aggregate(CupEquivalentPrice ~ Year + Form + Vegetable, data = Veggies, FUN = mean)


# Creates a summary statistic table for fruits
fruit_summary_form <- Fruits %>%
  group_by(Year, Form) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice))

# Creates a summary statistic table for vegetables
veggie_summary_form <- Veggies %>%
  group_by(Year, Form) %>%
  summarise(mean = mean(CupEquivalentPrice),
            median = median(CupEquivalentPrice),
            min = min(CupEquivalentPrice),
            max = max(CupEquivalentPrice))

# Views the fruit summary
cat("Fruit Summary:\n")
kable(fruit_summary_form, align = "c")

# Views the vegetable summary
cat("\nVeggie Summary:\n")
kable(veggie_summary_form, align = "c")

```

```{r}
plot_average_price_by_form <- function(data, title) {
  # Convert Year to factor
  data$Year <- as.factor(data$Year)
  
  p <- ggplot(data, aes(x = Year, y = mean, color = Form, group = Form)) +
    geom_line() +
    geom_point() +
    labs(title = title, x = "Year", y = "Retail Price (per equivalent cup, $)") +
    theme_minimal()
  
  return(p)
}

# Create plots for fruits and vegetables by form
plot_fruit_prices <- plot_average_price_by_form(fruit_summary_form, "Average Prices of Fruits by Form ")
plot_veggie_prices <- plot_average_price_by_form(veggie_summary_form, "Average Prices of Vegetables by Form")

# Display the plots
print(plot_fruit_prices)
print(plot_veggie_prices)

```

### 5. How many people can afford the recommended daily serving of fruits and vegetables across years?
```{r}
# Calculates 4% of per capita income
income_by_capita$eleven_percent <- income_by_capita$Value * 0.04

# Calculates 2% of per capita income
income_by_capita$four_percent =  income_by_capita$Value * 0.02

# Adjusts 'CupEquivalentPrice' to a yearly value by multiplying by 365
fruit_prices$CupEquivalentPrice_yearly <- fruit_prices$CupEquivalentPrice * 365
veggie_prices$CupEquivalentPrice_yearly <- veggie_prices$CupEquivalentPrice * 365

# Merges income_by_capita with fruit_prices by 'Year'
merged_fruits_2013 <- merge(subset(income_by_capita, Year == 2013), subset(fruit_prices, Year == 2013), by = "Year")
merged_fruits_Fruit2016 <- merge(subset(income_by_capita, Year == 2016), subset(fruit_prices, Year == 2016), by = "Year")
merged_fruits_Fruit2020 <- merge(subset(income_by_capita, Year == 2020), subset(fruit_prices, Year == 2020), by = "Year")

# Merges income_by_capita with veggie_prices by 'Year'
merged_veggies_2013 <- merge(subset(income_by_capita, Year == 2013), subset(veggie_prices, Year == 2013), by = "Year")
merged_veggies_Fruit2016 <- merge(subset(income_by_capita, Year == 2016), subset(veggie_prices, Year == 2016), by = "Year")
merged_veggies_Fruit2020 <- merge(subset(income_by_capita, Year == 2020), subset(veggie_prices, Year == 2020), by = "Year")

# Updates 'meetreq' column based on the condition for fruits
merged_fruits_2013$meetreq <- ifelse(merged_fruits_2013$four_percent > merged_fruits_2013$CupEquivalentPrice_yearly, "yes", "no")
merged_fruits_Fruit2016$meetreq <- ifelse(merged_fruits_Fruit2016$four_percent > merged_fruits_Fruit2016$CupEquivalentPrice_yearly, "yes", "no")
merged_fruits_Fruit2020$meetreq <- ifelse(merged_fruits_Fruit2020$four_percent > merged_fruits_Fruit2020$CupEquivalentPrice_yearly, "yes", "no")

# Updates 'meetreq' column based on the condition for vegetables
merged_veggies_2013$meetreq <- ifelse(merged_veggies_2013$four_percent > merged_veggies_2013$CupEquivalentPrice_yearly, "yes", "no")
merged_veggies_Fruit2016$meetreq <- ifelse(merged_veggies_Fruit2016$four_percent > merged_veggies_Fruit2016$CupEquivalentPrice_yearly, "yes", "no")
merged_veggies_Fruit2020$meetreq <- ifelse(merged_veggies_Fruit2020$four_percent > merged_veggies_Fruit2020$CupEquivalentPrice_yearly, "yes", "no")

# Creates comparison data frames for fruit in 2013
comparison_fruits_2013 <- data.frame(Year = merged_fruits_2013$Year,
                                     meetreq = merged_fruits_2013$meetreq,
                                     form = merged_fruits_2013$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_2013$CupEquivalentPrice_yearly)

# Creates comparison data frames for fruit in 2016
comparison_fruits_Fruit2016 <- data.frame(Year = merged_fruits_Fruit2016$Year,
                                     meetreq = merged_fruits_Fruit2016$meetreq,
                                     form = merged_fruits_Fruit2016$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_Fruit2016$CupEquivalentPrice_yearly)

# Creates comparison data frames for fruit in 2020
comparison_fruits_Fruit2020 <- data.frame(Year = merged_fruits_Fruit2020$Year,
                                     meetreq = merged_fruits_Fruit2020$meetreq,
                                     form = merged_fruits_Fruit2020$Form,
                                     CupEquivalentPrice_yearly = merged_fruits_Fruit2020$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2013
comparison_veggies_2013 <- data.frame(Year = merged_veggies_2013$Year,
                                      meetreq = merged_veggies_2013$meetreq,
                                      form = merged_veggies_2013$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_2013$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2016
comparison_veggies_Fruit2016 <- data.frame(Year = merged_veggies_Fruit2016$Year,
                                      meetreq = merged_veggies_Fruit2016$meetreq,
                                      form = merged_veggies_Fruit2016$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_Fruit2016$CupEquivalentPrice_yearly)

# Creates comparison data frames for vegetables in 2020
comparison_veggies_Fruit2020 <- data.frame(Year = merged_veggies_Fruit2020$Year,
                                      meetreq = merged_veggies_Fruit2020$meetreq,
                                      form = merged_veggies_Fruit2020$Form,
                                      CupEquivalentPrice_yearly = merged_veggies_Fruit2020$CupEquivalentPrice_yearly)


# Combines all comparison data frames for fruits and vegetables across different years
all_comparison_fruits <- rbind(comparison_fruits_2013, comparison_fruits_Fruit2016, comparison_fruits_Fruit2020)
all_comparison_veggies <- rbind(comparison_veggies_2013, comparison_veggies_Fruit2016, comparison_veggies_Fruit2020)

# Creates tables for fruits and vegetables
table_fruits <- table(all_comparison_fruits$meetreq, all_comparison_fruits$Year)
table_veggies <- table(all_comparison_veggies$meetreq, all_comparison_veggies$Year)

# Views fruit table
print("Fruits:")
print(table_fruits)

# Views vegetables table
print("Veggies:")
print(table_veggies)

```

```{r}
# Define the recommended daily serving cost for fruits and vegetables
recommended_daily_serving_cost <- 5  # Adjust this value if necessary

# Calculate the number of people who can afford the recommended daily serving
agg_income$people_afford <- agg_income$Value / recommended_daily_serving_cost

# Calculate the number of people who can't afford it
agg_income$people_cant_afford <- agg_income$Value - agg_income$people_afford

# Reshape the data to long format for easier plotting
agg_income_long <- pivot_longer(agg_income, cols = c(people_afford, people_cant_afford),
                                       names_to = "Affordability", values_to = "Number of People")

# Visualize
ggplot(agg_income_long, aes(x = as.factor(Year), y = `Number of People`, fill = factor(Affordability, levels = c("people_afford", "people_cant_afford")))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "People Affording Daily Fruit and Vegatable Servings",
       x = "Year", y = "Number of People") +
  scale_y_continuous(labels = scales::comma) +  
  scale_fill_manual(values = c("people_afford" = "skyblue", "people_cant_afford" = "coral"), 
                    labels = c("people_afford" = "Afford", "people_cant_afford" = "Can't Afford"),
                    name = "Affordability of Food Servings") +  # Setting legend title to NULL
  theme_minimal()
```
### 6. EXCURSION - How does the affordability of fruits and vegetables differ amongst their different forms?
```{r}
# Creates tables for fruits by form and year
table_fruits_form_2013 <- table(comparison_fruits_2013$meetreq, comparison_fruits_2013$form)
table_fruits_form_Fruit2016 <- table(comparison_fruits_Fruit2016$meetreq, comparison_fruits_Fruit2016$form)
table_fruits_form_Fruit2020 <- table(comparison_fruits_Fruit2020$meetreq, comparison_fruits_Fruit2020$form)

# Views 2013 fruit table
print("Fruits Summary by Form for 2013:")
print(table_fruits_form_2013)

# Views 2016 fruit table
print("Fruits Summary by Form for 2016:")
print(table_fruits_form_Fruit2016)

# Views 2020 fruit table
print("Fruits Summary by Form for 2020:")
print(table_fruits_form_Fruit2020)

# Convert tables to data frames
df_2013 <- as.data.frame(table_fruits_form_2013)
df_2016 <- as.data.frame(table_fruits_form_Fruit2016)
df_2020 <- as.data.frame(table_fruits_form_Fruit2020)

# Rename columns for clarity
colnames(df_2013) <- c("Affordability", "Form", "Count")
colnames(df_2016) <- c("Affordability", "Form", "Count")
colnames(df_2020) <- c("Affordability", "Form", "Count")

# Create plots for each year with a logarithmic scale for the y-axis to be better able to compare relative differences
plot_2013 <- ggplot(df_2013, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Fruit Affordability by Form (2013)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

plot_2016 <- ggplot(df_2016, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Fruit Affordability by Form (2016)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

plot_2020 <- ggplot(df_2020, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Fruit Affordability by Form (2020)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

# Display plots
plot_2013
plot_2016
plot_2020

#*we opted to show this graph collapsing year due to little variation between years()
# Combine data frames for all years
df_combined <- rbind(df_2013, df_2016, df_2020)

# Create a plot for all years combined
plot_combined <- ggplot(df_combined, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Fruit Affordability by Form",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

# Display the combined plot
plot_combined
```
```{r}
# Create tables for vegetables by form
table_veggies_form_2013 <- table(comparison_veggies_2013$meetreq, comparison_veggies_2013$form)
table_veggies_form_Fruit2016 <- table(comparison_veggies_Fruit2016$meetreq, comparison_veggies_Fruit2016$form)
table_veggies_form_Fruit2020 <- table(comparison_veggies_Fruit2020$meetreq, comparison_veggies_Fruit2020$form)

# Views 2013 vegetable table
print("Vegetable Summary by Form for 2013:")
print(table_veggies_form_2013)

# Views 2016 vegetable table
print("Vegetable Summary by Form for 2016:")
print(table_veggies_form_Fruit2016)

# Views 2020 vegetable table
print("Vegetable Summary by Form for 2020:")
print(table_veggies_form_Fruit2020)

# Convert tables to data frames
df_veggies_2013 <- as.data.frame(table_veggies_form_2013)
df_veggies_2016 <- as.data.frame(table_veggies_form_Fruit2016)
df_veggies_2020 <- as.data.frame(table_veggies_form_Fruit2020)

# Rename columns for clarity
colnames(df_veggies_2013) <- c("Affordability", "Form", "Count")
colnames(df_veggies_2016) <- c("Affordability", "Form", "Count")
colnames(df_veggies_2020) <- c("Affordability", "Form", "Count")

# Create plots for each year with a logarithmic scale for the y-axis
plot_veggies_2013 <- ggplot(df_veggies_2013, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Vegetable Affordability by Form (2013)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

plot_veggies_2016 <- ggplot(df_veggies_2016, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Vegetable Affordability by Form (2016)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

plot_veggies_2020 <- ggplot(df_veggies_2020, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Vegetable Affordability by Form (2020)",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

# Display plots
plot_veggies_2013
plot_veggies_2016
plot_veggies_2020

#*we opted to show this graph collapsing year due to little variation between years()
# Combine data frames for all years
df_combined_veg <- rbind(df_veggies_2013, df_veggies_2016, df_veggies_2020)

# Create a plot for all years combined
plot_combined_veg <- ggplot(df_combined_veg, aes(x = Form, y = Count, fill = Affordability)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Vegetable Affordability by Form",
       x = "Form",
       y = "Count (square root scale)",
       fill = "") + # Remove legend title
  scale_y_sqrt(labels = scales::comma_format()) + # Square root scale for y-axis
  scale_fill_manual(values = c("coral", "skyblue"),
                    labels = c("Can't Afford", "Afford")) + # Change legend labels
  theme_minimal()

# Display the combined plot
plot_combined_veg
```
